---
title: "Workshop Exercise: Joining and Pivoting"
author: "Jamison Ward" 
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---

# Introduction

In this workshop, you'll learn how to clean and analyze HIV and AIDS data across different countries and years.
We will start with raw datasets containing estimates of new and total HIV cases per country-year.
The goal is to clean and merge these datasets with population data.

Please refer to the lesson notes to aid you in the workshop exercise:

-   [Intro to Pivoting Data](https://the-graph-courses.github.io/data_untangled_wp/lessons/ls07_pivoting.html)
-   [Intro to Joining Datasets](https://the-graph-courses.github.io/fdar/FDAR_EN_joining_1/FDAR_EN_joining_1.html)

# Load Libraries

```{r}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, countrycode, esquisse)
```

# Data Import

We will work with two raw datasets, each containing estimates per country-year.
These data were accessed from the Gapminder foundation, at [www.gapminder.org/data](https://www.gapminder.org/data).

1.  **People living with HIV:** This dataset contains the total number of people currently infected with HIV (*data/hiv_prevalence.csv*).
2.  **New HIV infections:** This dataset provides the total number of people newly infected with HIV during a given year (*data/hiv_incidence.csv*).

Load the datasets into R using `read_csv()`.

```{r}
# Load raw data from specified paths
total_hiv_raw <- read_csv(here("data/hiv_prevalence.csv"))
new_hiv_raw <- read_csv(here("data/hiv_incidence.csv"))
```

# Explore the Datasets

Use functions like `head()`, `View()`, and `summary()` to inspect the datasets.

```{r}
# view(total_hiv_raw)
# view(new_hiv_raw)
# summary(total_hiv_raw)
# summary(new_hiv_raw)
```

# Data Pivoting

Now that we have imported the data it's time to pivot it.

The datasets are currently in wide format, with each year as a separate column, something like this:

| country     | 1990  | 1991  | 1992  | 1993  | 1994  |
|-------------|-------|-------|-------|-------|-------|
| Afghanistan | 600   | 600   | 1000  | 1100  | 1200  |
| Angola      | 33000 | 39000 | 46000 | 54000 | 62000 |
| Argentina   | 27000 | 31000 | 35000 | 39000 | 43000 |
| Armenia     | 350   | 600   | 600   | 600   | 1000  |
| Australia   | 12000 | 12000 | 12000 | 11000 | 11000 |
| Austria     | 600   | 600   | 1000  | 1300  | 1500  |

We need to convert this to long format, where each row represents a single year-country combination.
For example:

| country     | year | cases |
|-------------|------|-------|
| Afghanistan | 1990 | 600   |
| Afghanistan | 1991 | 600   |
| Afghanistan | 1992 | 1000  |
| Afghanistan | 1993 | 1100  |
| Afghanistan | 1994 | 1200  |

## Instructions for Pivoting `new_hiv_raw` Data

Reshape the `new_hiv_raw` dataset by converting the columns that represent different years into a new column called `year`, and place the corresponding values into a new column called `new_cases`.

**HINTS**:

1.  Use `pivot_longer()` on `new_hiv_raw`.
2.  Exclude the `country` and `country_code` columns from being pivoted by specifying `cols = -c(country, country_code)`.
3.  Set the names of the new columns inside `pivot_longer()` using:
    -   `names_to = "year"` to label the column that will store the years.
    -   `values_to = "new_cases"` to store the values representing new HIV cases under the `new_cases` column.

```{r}
new_hiv_long <- new_hiv_raw %>% 
  pivot_longer(
    cols = `2000`:`2011`,
    names_to = "year",
    values_to = "new_cases"
    )

new_hiv_long
```

------------------------------------------------------------------------

## Instructions for Pivoting `total_hiv_raw` Data

In a similar manner, pivot the `total_hiv_raw` dataset to a long format.
Convert the year columns into a new column called `year` and the counts into `total_cases`.

```{r}
total_hiv_long <- total_hiv_raw %>% 
  pivot_longer(
    cols = `2000`:`2011`,
    names_to = "year",
    values_to = "total_cases"
    )

total_hiv_long
```

**CHECKPOINT:**\
After performing these steps, make sure that you have two reshaped datasets:

-   `new_hiv_long` with **1704 rows and 4 columns**: `country`, `country_code`, `year`, and `new_cases`.

-   `total_hiv_long` with **1776 rows and 4 columns**: `country`, `country_code`, `year`, and `total_cases`.

# Joining the HIV Datasets

Now that both datasets are in long format, we can join them.
Note that you need to use the `country`, `country_code`, and `year` columns as the keys (`by = c("country", "country_code", "year")`).

Use `full_join()` so that we can keep as much information as possible.

```{r}
hiv_data_combined <- full_join(
  new_hiv_long, total_hiv_long, by = c("country", "country_code", "year")
)

hiv_data_combined
```

**CHECKPOINT:** The combined dataset should have columns for `country`, `country_code`, `year`, `total_cases`, and `new_cases`.

# Add Population Data

The `world_bank_pop` dataset from **tidyr** contains information on population per year from 2000 to 2017.

```{r}
tidyr::world_bank_pop # Run this code to see the dataset
# view(tidyr::world_bank_pop)
```

Note that there are several measures of population described in the `indicator` column.
We are interested in the total population of each country, which is the `SP.POP.TOTL` indicator.

Additionally, note that the `country` column doesn't have full country names, but the 3-letter country codes.
This will make it easier to join with our HIV dataset, since country codes are consistent across datasets (unlike full country names).

## Filter and Pivot Population Data

Next, complete the following transformations on the `world_bank_pop` dataset:

-   Filter to just the `SP.POP.TOTL` indicator

-   Pivot the data to a long format.

-   Rename the "`country`" column to "`country_code`" so that we can join it properly later on.

-   Remove the `indicator` column.

-   Store the result as `population_long`.

```{r}
population_long <- tidyr::world_bank_pop %>% 
  filter(indicator == "SP.POP.TOTL") %>% 
  pivot_longer(
    cols = `2000`:`2017`,
    names_to = "year",
    values_to = "total_population") %>% 
  rename(country_code = country) %>% 
  select(-indicator)

population_long
```

**CHECKPOINT:**\
After performing these steps, make sure that you have a dataset named `population_long` with **4788 rows and 3 columns**: `country_code`, `year`, and `population`.

## Join Population Data with HIV Data

Use `left_join()` to add the population information to your HIV data combined dataset (the HIV data should be the left or primary dataset).
You should use both "country_code" and "year" as keys.

```{r}
# Your code here
hiv_data_combined_pop <- left_join(
  hiv_data_combined, population_long, by = c("country_code", "year")
)
  
hiv_data_combined_pop
```

**CHECKPOINT:** Verify that `hiv_data_combined_pop` now contains a `population` column.

## Calculate Incidence Rate and Prevalence Rate

Lastly, calculate the incidence rate and prevalence rate per 100,000 people:

-   **Incidence Rate** = New Cases / Population \* 100000
-   **Prevalence Rate** = Total Cases / Population \* 100000

```{r}
hiv_data_final <- hiv_data_combined_pop %>% 
  mutate(incidence_rate = new_cases / total_population * 100000,
         prevalence_rate = total_cases / total_population * 100000
         )

hiv_data_final
```

**CHECKPOINT:** `hiv_data_final` should have **1776 rows and 8 columns**: `country`, `country_code`, `year`, `total_cases`, `new_cases`, `population`, `incidence_rate` and `prevalence_rate` (the names don't need to be exact matches as long as the data is equivalent).

# Plot Incidence Rates

Now, you'll select 3 countries of your choice and plot their incidence rate over time.

Before passing your data into `esquisser()`, you should:

-   Filter the data to narrow down the countries. Use the syntax `filter(country %in% c("COUNTRY1", "COUNTRY2", "COUNTRY3")`.
-   Convert the `year` column to a `numeric` type with `mutate()` and `as.numeric()`.

**HINT**: Some countries have a many years with missing data, especially for `new_cases`.
View and check your `hiv_data_combined_pop` dataset to ensure that your chosen countries are not full of NAs.

```{r}
# Create data subset with 3 countries
hiv_data_mini <- hiv_data_final %>%
  filter(country_code %in% c("IND", "IDN", "MEX")) %>% 
  mutate(year = as.numeric(year))

hiv_data_mini
```

Now, use `hiv_data_mini` and plot incidence rates over time.

**HINT**: Put `year` on the x-axis, `incidence_rate` on the y-axis, and color by `country`.
Use a line graph for best visualisation.

```{r}
hiv_indicence_rate_plot <- ggplot(hiv_data_mini) +
  aes(x = year, y = incidence_rate, colour = country) +
  geom_line() +
  scale_color_hue(direction = 1) +
  scale_x_continuous(breaks = c(2000:2011)) +
  scale_y_continuous(breaks = seq(0,30,5), limits = c(0,30)) +
  labs(
    x = "Year",
    y = "Incidence Rate",
    title = "HIV Incidence Rates",
    caption = "Incidence rates are defined as the number of new HIV cases per 100,
   000 individuals."
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 18L, hjust = 0.5))

ggsave(hiv_indicence_rate_plot, filename = here("outputs/hiv_indicence_rate_plot.png"))
```

# Plot Prevalence Rates

Again, use `hiv_data_mini` and plot their prevalence rates over time, using a line graph.

```{r}
hiv_prevalence_rate_plot <- ggplot(hiv_data_mini) +
  aes(x = year, y = prevalence_rate, colour = country) +
  geom_line() +
  scale_color_hue(direction = 1) +
  scale_x_continuous(breaks = c(2000:2011)) +
  scale_y_continuous(breaks = seq(0,250,50), limits = c(0,250)) +
  labs(
    x = "Year",
    y = "Prevalence Rate",
    title = "HIV Prevalence Rates",
    caption = "Prevalence rates are defined as the number of existing HIV cases per 100,
   000 individuals."
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 18L, hjust = 0.5))

ggsave(hiv_prevalence_rate_plot, filename = here("outputs/hiv_prevalence_rate_plot.png"))
```

# Optional Ungraded Challenge

In this challenge, you will:

1.  **Add Regional Information:** Use the `countrycode` package to get regional (continent) information for each country.
2.  **Merge Continent Data:** Merge the continent data with your dataset.
3.  **Identify Highest Prevalence Rates per Continent:** Find the countries with the highest prevalence rate in 2011 for each continent.
4.  **Create a Bar Chart:** Make a bar chart comparing the 2011 prevalence rates of these countries.

```{r}
# view(countrycode::codelist)

country_info <- countrycode::codelist %>%  # Parse countrycode dataset for country codes
 select(iso3c, continent) %>% 
  rename(country_code = iso3c)

hiv_data_final_continent <- left_join(  # Add continent data to the HIV dataset
  hiv_data_final, country_info, by = "country_code"
  )

# sum(is.na(hiv_data_final_continent$continent))  # Check to see if each country was assigned to a continent

hiv_top_countries <- hiv_data_final_continent %>%  # Retrieve country with highest HIV incidence in 2011 for each continent
  group_by(continent) %>% 
  filter(year == 2011) %>% 
  filter(prevalence_rate == max(prevalence_rate, na.rm = TRUE))

hiv_top_countries_plot <- ggplot(hiv_top_countries) +
  aes(x = country, y = prevalence_rate, fill = continent) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  scale_y_continuous(breaks = seq(0,18000,2000), limits = c(0,18000)) +
  labs(
    x = "Country",
    y = "Prevalence Rate",
    title = "Top HIV Prevalence Rates in 2011",
    caption = "Prevalence rates are defined as the number of existing HIV cases per 100,
   000 individuals."
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 18L, hjust = 0.5))

ggsave(hiv_top_countries_plot, filename = here("outputs/hiv_top_countries_plot.png"))


hiv_top_countries_plot_log <- ggplot(hiv_top_countries) +
  aes(x = country, y = prevalence_rate, fill = continent) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  scale_y_continuous(trans = "log10") +
  labs(
    x = "Country",
    y = "Prevalence Rate",
    title = "Top HIV Prevalence Rates in 2011",
    caption = "Prevalence rates are defined as the number of existing HIV cases per 100,
   000 individuals. Note the log y-axis scale."
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 18L, hjust = 0.5))

ggsave(hiv_top_countries_plot_log, filename = here("outputs/hiv_top_countries_plot_log.png"))
```
